Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-10-29 paper.tex
fixes

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\uir}{\msf{U^{IR}}}
\newcommand{\elir}{\msf{El^{IR}}}
\newcommand{\ult}{\U_{<}}
\newcommand{\mkMor}{\msf{mk}\!_<}
\newcommand{\unMor}{\msf{un}\!_<}
\newcommand{\mkLvl}{\msf{mk}_{\Lvl}}
\newcommand{\unLvl}{\msf{un}_{\Lvl}}

@@

In the following, whenever we introduce a type or term former, we always assume
that it is natural with respect to substitution, i.e.\ all type and term formers
have a corresponding substitution rule.

@@

\begin{proof}
From the $\ext$-preservation isomorphism and the defining isomorphisms of
comprehension, we get $\q' : \Tm\,(\Gamma \ext F\,A)\,(A[\p])$ such that $F\,\q'
= \q$ and $\q'[\p,\,F\,\q] = \q$. Now, for $t : \Tm\,\Gamma\,(F\,A)$, we
define $F^{-1}\,t$ as $\q'[\id,\,t] : \Tm\,\Gamma\,A$. We get the following:
\begin{alignat*}{3}
  & F(F^{-1}\,t) = F(\q'[\id,\,t]) = (F\,\q')[\id,\,t] = \q'[\id,\,t] = t \\
  & F^{-1}(F\,t) = \q'[\id,\,F\,t] = \q'[\id,\,(F\,\q)[\id,\,\,t]] = \q'[\p,\,F\,\q][\id,\,t]
    = \q[\id,\,t] = t
\end{alignat*}
More concisely, $F$ is invertible on the generic term $\q$, which implies invertibility
on any term.
\end{proof}

@@

Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a morphism between them. This
corresponds to a basic version of \emph{two-level type theory} \cite{twolevel}. This theory can be
interpreted as having ($\Ty_1$, $\Tm_1$) as a metaprogramming layer which can generate object-level
programs in the ($\Ty_0$, $\Tm_0$) layer. Lifted types correspond to types of object-level
expressions; for example, $\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
$\Lift\,\Bool_0$ is the meta-level type of $\Bool_0$-expressions, and $\Bool_1 : \Ty_1\,\Gamma$ is
the type of meta-level Booleans. It is possible to compute a $\Bool_0$ from a $\Bool_1$. Given $b :
\Tm_1\,\Gamma\,\Bool_1$, we can construct
$\down(\mathsf{if}\,b\,\mathsf{then}\up\!\true_0\,\mathsf{else}\up\!\false_0) :
\Tm_0\,\Gamma\,\Bool_0$.

But there is no way to compute a $\Bool_1$ from a $\Bool_0$: we can try to lift
the input, but there is no elimination rule for $\Lift\,\Bool_0$ in
$\Ty_1$. Conceptually, terms of $\Lift\,\Bool_0$ are \emph{expressions}, so they
are not necessarily $\true$ or $\false$, they can also be variables or neutral
expressions, so the usual Boolean elimination is not justified for them. In
general, the setup ensures that we can map from positive types in $\Ty_1$
to their counterparts in $\Ty_0$, but not the other way around, while negative
types are preserved by $\Lift$ up to type isomorphism.

The rules of two-level type theory appear to model a form of generative two-stage compilation with
dependent types. Comparing this system to e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can
relate $\Lift$ to $\mathsf{code}$, $\up$ to the quasi-quotation operation $<\!\blank\!>$, and
$\down$ to escape $\sim\!\blank$.

While the staging aspect could be subject of future research, we currently focus
on ``sizing'' hierarchies instead of staging hierarchies. This means that we
want to eliminate from any family to any other family which is connected by
a morphism.

@@

\noindent Some examples for preservation equations for type and term formers:
\begin{alignat*}{3}
  & \Lift\,(\Pi\,(x : A) B)   && =\,\,\,\,&& \Pi\,(x : \Lift\,A)(\Lift\, (B[x \mapsto\,\down\!x]))\\
  & \up(\lambda\,(x : A).\,t) && =&& \lambda\,(x : \Lift\,A).\up(t[x\mapsto\down\!x])\\
  & \Lift\,\Bool_0            && =&& \Bool_1\\
  & \up\true_0                && =&& \true_1
\end{alignat*}
In general, we can skip specifying preservation for $\down$, since it follows
from $\up$ preservation equations.

Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_1$). Now, we can
eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
\Tm_1\,\Gamma\,(\Lift\,\Bool_0)$, hence $\up\!b :
\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
computes appropriately on canonical terms: if $b$ is $\true_0$, we get
$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.

@@

\begin{mydefinition}\label{def:strict_inclusion}
A \emph{strict family inclusion} between ($\Ty_0$, $\Tm_0$) and ($\Ty_1$,
$\Tm_1$) is a family inclusion ($\Lift$, $\up$, $\down$) for which the following equations hold:
\begin{alignat}{2}
  & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
  & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
  & \up\!t &&= t                                    \label{eq:cumlift}
\end{alignat}
\end{mydefinition}

A strict inclusion corresponds to Sterling's \emph{algebraic cumulativity}
\cite{sterling2019algebraic}. The additional equations are a matter of
convenience: they allow us to omit term liftings in informal syntax\footnote{In
  a proof assistant, often we would still have to explicitly transport along the
  strict inclusion equations.}. Most of the time we can also omit level
annotations on term formers. For example, we have $\true_0 :
\Tm_0\,\Gamma\,\Bool_0$, but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$,
hence $\true_0 : \Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally
equal to $\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using
simply $\true$ is fine whenever the family is clear from context.

The definitional equality of $\true_0$ and $\true_1$ is important; without it
canonicity would fail, since $\true_0$, $\false_0$, $\true_1$ and $\false_1$
would be four definitionally distinct inhabitants of $\Bool_1$. See Luo
\cite{luo2012notes} for a discussion of related issues with cumulativity. It is
not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$
together with equation \ref{eq:cumlift} to identify term formers in different
families. The other direction $\down\!t = t$ is immediately derivable.

@@

We would like to describe a range of setups with multiple families and morphisms
between them. In this subsection we describe the indexing structures for such
family diagrams. First, some preliminary definitions.

\begin{mydefinition} The \emph{accessibility predicate} on relations is defined by the following
inductive rules:
\begin{alignat*}{3}
  & \Acc : \{A : \Seti\}\to(R : A \to A \to \Seti) \to A \to \Seti \\
  & \acc : \{a : A\} \to ((\mi{a'} : A) \to R\,\mi{a'}\,a \to \Acc\,R\,\mi{a'}) \to \Acc\,R\,a
\end{alignat*}
\end{mydefinition}

\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$, all
descending $R$-chains must be finite. This is ensured by the universal property
of the inductive definition, as all inductive types must be well-founded.

\begin{mylemma}\label{lem:accprop}
Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
equal \cite[Lemma 10.3.4]{hottbook}.
\end{mylemma}

@@

Our choice of level structures and diagrams is motivated by the following.  First, we do not need
identity morphisms in levels, because they would be mapped to trivial liftings, which are not
interesting in our setting. Second, we do not need proof-relevant level morphisms, since any
parallel pair of morphisms gives rise to isomorphic types. Concretely, given $p : i < j$ and $q : i
< j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq \Tm_i\,\Gamma\,A \simeq
\Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$ and $\Lift\,q\,A$ are in the same family, we
can internally prove them isomorphic using function types and identity types. That said, every
construction in this paper would still work with inverse categories as level structures.

@@

At this point, we can talk about family diagrams, but no previously seen type
former depends on levels in an interesting way. For example, $\Bool_i$ has the
same canonical inhabitants as $\Bool_j$, for any $i$ and $j$. Universes
introduce dependency on levels, by serving as classifiers for smaller families
internally to larger families.

\begin{mydefinition} A family diagram supports \emph{universe formation} if it supports the following:
\begin{alignat*}{3}
  & \U             &&: (i\,j : \Lvl) \to i < j \to \Ty_j\,\Gamma\\
  & \mathsf{LiftU} &&: \Lift_{j}^{k}\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)
\end{alignat*}
\end{mydefinition}
We also need a way to pin down universes as classifiers. We consider two variants.
\begin{mydefinition}
A family diagram has \emph{Coquand universes} if it has universe formation and
additionally supports $\El : \Tm_j\,\Gamma\,(\U\,i\,j\,p) \to \Ty_i\,\Gamma$, and
its inverse $\Code : \Ty_i\,\Gamma \to \Tm_j\,\Gamma\,(\U\,i\,j\,p)$.
\end{mydefinition}
\begin{mydefinition}
A family diagram has \emph{Russell universes} if it has Coquand universes and
additionally satisfies $\Tm_j\,\Gamma\,(\U\,i\,j\,p) = \Ty_i\,\Gamma$ and
$\El\,t = t$.
\end{mydefinition}

@@

In this section we give a model for a type theory with generalized
universes. Let's make the notion of model concrete first.

@@

The task is to interpret the $\Lvl$-many universes of TTGU using a metatheoretic
feature which is more compact and already well-studied. For this, we need to
define a $\Lvl$-indexed type of type codes. Since $\Lvl$ and $\!\blank<\!\blank$
can be arbitrary, we effectively need to define transfinite hierarchies of
codes. We use an inductive-recursive (IR) \cite{dybjer99finite} definition, for
the following reasons.

First, IR is already supported in the proof assistant Agda, and it is very
useful to be able to sketch out ideas in a machine-checked setting. It would be
much harder to do the same when developing semantics in set theory.

Second, could we use simpler type-theoretic features, like super universes
\cite{Palmgren98onuniverses} or Mahlo universes \cite{setzer00mahlo}? These are
sufficient to model transfinite hierarchies. However, it is not clear how to
additionally support \emph{recursive sub-universes}, which in our work corresponds to
the type former preservation property of $\Lift$. Thus, we give a custom IR
definition which corresponds more directly to TTGU structure.

\begin{mydefinition}[IR codes]
  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$
  and $\elir$ by simultaneous IR:
\begin{alignat*}{5}
  & \uir   &&: \Set0
     && \rlap{$\elir : \uir \to \Set0$} \\
  & \U'    &&: (j : \Lvl) \to j < i \to \uir
     && \elir\,(\U'\,j\,p) &&= f\,j\,p \\
  & \Pi'   &&: (A : \uir) \to (\elir\,A \to \uir) \to \uir \hspace{2em}
     && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
  & \bot' &&: \uir
     && \elir\,\bot'      &&= \bot\\
  & \Bool' &&: \uir
     && \elir\,\Bool'     &&= \Bool
\end{alignat*}
\end{mydefinition}
We use the prime accents ($'$) to disambiguate IR codes from type formers in TTGU or the metatheory.
For basic type formers, we only include codes for function types, the empty type, and $\Bool$.
Other type formers are straightforward to add (and we do have more in the Agda formalization).

@@

\subsection{IR Model of TTGU}

@@

For any definition of internal levels, we may extend the specification of TTDL
with the corresponding reflection rule, thereby getting an algebraic signature
for a type theory with first-class levels (TTFL). We can easily get a TTFL with
an IR model in the following way. First, pick an external $\Lvl$ structure which
a) satisfies the bootstrapping assumption b) has sets of levels and morphisms
which can be represented with syntactic TTDL types.

@@

Note that $\Pi$ is contravariant in the domain. This is easily supported
with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
\cite{timany18cumulative}, where function domains are invariant.

@@

\item If levels are given by a limit ordinal, then every TTFL type is contained
in a universe. If levels form a successor ordinal, then this is not the
case. For example, Agda 2.6.1 has $\omega + 1$ levels (externally), where
$\Set{\omega}$ is the topmost universe, but $\Set{\omega}$ has no type.

@@

There have been proposals for strengthening universes with various closure
principles and universe operators. Palmgren's super universes and higher-order
universes \cite{Palmgren98onuniverses} and Setzer's Mahlo universes
\cite{setzer00mahlo} are examples for this. These are sufficient to model
transfinite hierarchies, but as we noted in Section
\ref{sec:inductive_recursive_codes}, we do not know how to model strict
inclusions with them. Variants of induction-recursion \cite{dybjer99finite,
  indexedir, positiveir} are particularly flexible and powerful extensions to
universes.

@@

In the current work, we developed a framework for modeling a variety of universe
features in type theories. At this point, we may ask the question: if IR is
sufficient to model every feature, why not simply add IR support in a practical
implementation, and drop the menagerie of universe features?

The answer is that IR provides is a \emph{deep embedding} of universe features,
which is usually less convenient to use than \emph{native} features. For example,
both Coq and Agda have powerful automatic solving for filling out implicit
universe levels. We also do not have to invoke $\El$ or the $\ult$ computation
rule explicitly, and in Coq we can use implicit syntax for subtyping instead of
explicit coercions.

@@

First, besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. We did keep
canonicity in mind when specifying the systems in this paper. Hopefully the
usual proof method of gluing (in other words, proof-relevant logical predicates)
\cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
this paper.

@@

Lastly, we do not handle impredicative universes. The main reason for this is
that we do not know the consistency of having IR and impredicative function
space together in the same universe, and modeling impredicativity seems to
require this assumption in the metatheory. This could be investigated as well
in future work.

>>>>

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{{\uparrow}}
\newcommand{\down}{{\downarrow}}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\uir}{\msf{U^{IR}}}
\newcommand{\elir}{\msf{El^{IR}}}
\newcommand{\ult}{\U_{<}}
\newcommand{\mkMor}{\msf{mk}\!_<}
\newcommand{\unMor}{\msf{un}\!_<}
\newcommand{\mkLvl}{\msf{mk}_{\Lvl}}
\newcommand{\unLvl}{\msf{un}_{\Lvl}}

@@

In the following, whenever we introduce a type or term former, we always assume
that it is natural with respect to substitution, i.e.\ all type and term formers
have a corresponding substitution rule. This convention could be made precise by
working in a framework for higher-order abstract syntax, where all specified
structure is automatically stable under substitution
\cite{cubicalnorm,uemura,bocquet2021relative}. While this can be effective at
reducing formal clutter, this paper only presents models which are technically
straightforward, so we choose not to use higher-order signatures, in order to
make the presentation more direct.

@@

\begin{proof}
From the $\ext$-preservation isomorphism and the defining isomorphisms of
comprehension, we get $\q' : \Tm\,(\Gamma \ext F\,A)\,(A[\p])$ such that $F\,\q'
= \q$ and $\q'[\p,\,F\,\q] = \q$. Now, for $t : \Tm\,\Gamma\,(F\,A)$, we
define $F^{-1}\,t$ as $\q'[\id,\,t] : \Tm\,\Gamma\,A$. We get the following:
\begin{alignat*}{3}
  & F(F^{-1}\,t) = F(\q'[\id,\,t]) = (F\,\q')[\id,\,t] = \q[\id,\,t] = t \\
  & F^{-1}(F\,t) = \q'[\id,\,F\,t] = \q'[\id,\,(F\,\q)[\id,\,\,t]] = \q'[\p,\,F\,\q][\id,\,t]
    = \q[\id,\,t] = t
\end{alignat*}
More concisely, $F$ is invertible on the generic term $\q$, which implies invertibility
on any term.
\end{proof}

@@

Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
morphism between them. This corresponds to a basic version of \emph{two-level
type theory} \cite{twolevel}. This theory has an interpretation in presheaves
over the category of contexts of some chosen model of a type theory, where
($\Ty_0$, $\Tm_0$) is modeled using structure in the chosen model, and ($\Ty_1$,
$\Tm_1$) is modeled using presheaf constructions. More illustratively, this
means interpreting ($\Ty_1$, $\Tm_1$) as a metaprogramming layer which can
generate object-level constructions in the ($\Ty_0$, $\Tm_0$) layer. Lifted
types correspond to types of object-level terms; for example, $\Bool_0 :
\Ty_0\,\Gamma$ is the object-level type of Booleans, while $\Lift\,\Bool_0$ is
the meta-level type of $\Bool_0$-terms, and $\Bool_1 : \Ty_1\,\Gamma$ is the
type of meta-level Booleans. It is possible to compute a $\Bool_0$ from a
$\Bool_1$. Given $b : \Tm_1\,\Gamma\,\Bool_1$, we can construct
$\down(\mathsf{if}\,b\,\mathsf{then}\,\up\true_0\,\mathsf{else}\,\up\false_0) :
\Tm_0\,\Gamma\,\Bool_0$. But there is no way to compute a $\Bool_1$ from a
$\Bool_0$: we can try to lift the input, but there is no elimination rule for
$\Lift\,\Bool_0$ in $\Ty_1$.

Hence, plain family morphisms can model a metaprogramming hierarchy, but
currently we are aiming for ``sizing'' hierarchies instead. This means that we
want to eliminate from any family to any other family which is connected by a
morphism.

@@

\noindent Some examples for preservation equations for type and term formers:
\begin{alignat*}{3}
  & \Lift\,(\Pi\,(x : A) B)   && =\,\,\,\,&& \Pi\,(x : \Lift\,A)(\Lift\, (B[x \mapsto\,\down x]))\\
  & \up(\lambda\,(x : A).\,t) && =&& \lambda\,(x : \Lift\,A).\,\up(t[x\mapsto\down x])\\
  & \Lift\,\Bool_0            && =&& \Bool_1\\
  & \up\true_0                && =&& \true_1
\end{alignat*}
In general, we can skip specifying preservation for $\down$, since it follows
from $\up$ preservation equations.

Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_1$). Now, we can
eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
\Tm_0\,\Gamma\,\Bool_0$, we also have $\up b :
\Tm_1\,\Gamma\,(\Lift\,\Bool_0)$, hence $\up b :
\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
$\mathsf{if}\,\up b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
computes appropriately on canonical terms: if $b$ is $\true_0$, we get
$\up \true_0 = \true_1$ as the if-then-else scrutinee.

@@

\begin{mydefinition}\label{def:strict_inclusion}
A \emph{strict family inclusion} between ($\Ty_0$, $\Tm_0$) and ($\Ty_1$,
$\Tm_1$) is a family inclusion ($\Lift$, $\up$, $\down$) for which the following equations hold:
\begin{alignat}{2}
  & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
  & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
  & \up t &&= t                                    \label{eq:cumlift}
\end{alignat}
\end{mydefinition}

A strict inclusion corresponds to Sterling's \emph{algebraic cumulativity}
\cite{sterling2019algebraic}. The additional equations are a matter of
convenience: they allow us to omit term liftings in informal syntax\footnote{In
  a proof assistant, often we would still have to explicitly transport along the
  strict inclusion equations.}. Most of the time we can also omit level
annotations on term formers. For example, we have $\true_0 :
\Tm_0\,\Gamma\,\Bool_0$, but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$,
hence $\true_0 : \Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally
equal to $\true_1$, since $\true_0 =\,\up \true_0 = \true_1$. Thus, using
simply $\true$ is fine whenever the family is clear from context.

The definitional equality of $\true_0$ and $\true_1$ is important; without it
canonicity would fail, since $\true_0$, $\false_0$, $\true_1$ and $\false_1$
would be four definitionally distinct inhabitants of $\Bool_1$. See Luo
\cite{luo2012notes} for a discussion of related issues with cumulativity. It is
not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}. We need $\up$
together with equation \ref{eq:cumlift} to identify term formers in different
families. The other direction $\down t = t$ is immediately derivable.

@@

We would like to describe a range of setups with multiple families and morphisms
between them. In this subsection we describe the indexing structures for such
family diagrams. First, we specify a notion of well-foundedness, which will be
used to preclude size paradoxes in universe hierarchies.

\begin{mydefinition}
The \emph{accessibility predicate} on relations is defined by the following
inductive rules:
\begin{alignat*}{3}
  & \Acc : \{A : \Seti\}\to(R : A \to A \to \Seti) \to A \to \Seti \\
  & \acc : \{a : A\} \to ((\mi{a'} : A) \to R\,\mi{a'}\,a \to \Acc\,R\,\mi{a'}) \to \Acc\,R\,a
\end{alignat*}
\end{mydefinition}
\noindent See \cite{aczel1977introduction} and \cite[Section~10.3]{hottbook} for
further exposition. An inhabitant of $\Acc\,R\,a$ proves that starting from $a :
A$, all descending $R$-chains must be finite. This is ensured by the universal
property of the inductive definition.

\begin{mylemma}\label{lem:accprop}
All inhabitants of $\Acc\,R\,a$ are equal \cite[Lemma 10.3.4]{hottbook}. In
other words, accessibility is proof-irrelevant.
\end{mylemma}

@@

Our choice of level structures and diagrams is motivated by the following.  First, we do not need
identity morphisms in levels, because they would be mapped to trivial liftings, which are not
interesting in our setting. Second, we do not need proof-relevant level morphisms, since any
parallel pair of morphisms gives rise to isomorphic types. Concretely, given $p : i < j$ and $q : i
< j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq \Tm_i\,\Gamma\,A \simeq
\Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$ and $\Lift\,q\,A$ are in the same family, we
can internally prove them isomorphic using function types and identity types. That said, every
construction in this paper would still work with direct categories as level structures.

@@

At this point, we can talk about family diagrams, but no previously seen type
former depends on levels in an interesting way. For example, $\Bool_i$ has the
same inhabitants as $\Bool_j$, for any $i$ and $j$. Universes introduce
dependency on levels, by serving as classifiers for smaller families internally
to larger families.

\begin{mydefinition} A family diagram supports \emph{universe formation} if it supports the following:
\begin{alignat*}{3}
  & \U             &&: (i\,j : \Lvl) \to i < j \to \Ty_j\,\Gamma\\
  & \mathsf{LiftU} &&: \Lift_{j}^{k}\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)
\end{alignat*}
\end{mydefinition}
We also need a way to pin down universes as classifiers. We consider two variants.
\begin{mydefinition}
A family diagram has \emph{Coquand universes} \cite{coquandnorm} if it has universe formation and
additionally supports $\El : \Tm_j\,\Gamma\,(\U\,i\,j\,p) \to \Ty_i\,\Gamma$, and
its inverse $\Code : \Ty_i\,\Gamma \to \Tm_j\,\Gamma\,(\U\,i\,j\,p)$.
\end{mydefinition}
\begin{mydefinition}
A family diagram has \emph{Russell universes} if it has Coquand universes and
additionally satisfies $\Tm_j\,\Gamma\,(\U\,i\,j\,p) = \Ty_i\,\Gamma$ and
$\El\,t = t$.
\end{mydefinition}

@@

In this section we give a model for a type theory with generalized
universes. Let us make the notion of model concrete first.

@@

The task is to interpret the $\Lvl$-many universes of TTGU using an assumed
metatheoretic feature. For this, we need to define a $\Lvl$-indexed type of type
codes. Since $\Lvl$ and $\!\blank<\!\blank$ can be arbitrary, we effectively
need to define transfinite hierarchies of codes. We use an inductive-recursive
\cite{dybjer99finite} definition for the following reasons.

First, induction-recursion is already supported in the Agda proof assistant, and
it is very useful to be able to sketch out ideas in a machine-checked
setting. It would be much harder to do the same when developing semantics in set
theory.

Second, could we use type-theoretic features with simpler specifications than
induction-recursion, such as super universes \cite{Palmgren98onuniverses} or
Mahlo universes \cite{setzer00mahlo}? These are sufficient to model transfinite
hierarchies. However, using these it is not clear how to additionally support
the strict type former preservation property of $\Lift$\footnote{Palmgren calls
this property as having \emph{recursive sub-universes}
\cite{Palmgren98onuniverses}.}.

Therefore, we give a custom definition using induction-recursion, which
corresponds more directly to TTGU structure. Our definition is essentially the
same as McBride's redundancy-free hierarchy in
\cite[Section~6.3.1]{mcbride2015datatypes}, but we generalize levels from
natural numbers to arbitrary level structures.

\begin{mydefinition}[Codes for the universe]
Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$ and
$\elir$ by induction-recursion:
\begin{alignat*}{5}
  & \uir   &&: \Set0
     && \rlap{$\elir : \uir \to \Set0$} \\
  & \U'    &&: (j : \Lvl) \to j < i \to \uir
     && \elir\,(\U'\,j\,p) &&= f\,j\,p \\
  & \Pi'   &&: (A : \uir) \to (\elir\,A \to \uir) \to \uir \hspace{2em}
     && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
  & \bot' &&: \uir
     && \elir\,\bot'      &&= \bot\\
  & \Bool' &&: \uir
     && \elir\,\Bool'     &&= \Bool
\end{alignat*}
\end{mydefinition}
We use the prime accents ($'$) to disambiguate inductive-recursive codes from
type formers in TTGU or the metatheory. For basic type formers, we only include
codes for function types, the empty type, and $\Bool$. Other type formers are
straightforward to add (and we do have more in the Agda formalization).

@@

\subsection{Inductive-Recursive Model of TTGU}

@@

For any definition of internal levels, we may extend the specification of TTDL
with the corresponding reflection rule, thereby getting an algebraic signature
for a type theory with first-class levels (TTFL). We can easily get a TTFL with
an inductive-recursive model in the following way. First, we pick an external
$\Lvl$ structure which a) satisfies the bootstrapping assumption b) has sets of
levels and morphisms which can be represented with syntactic TTDL types.

@@

Note that $\Pi$ is contravariant in the domain. This is easily supported with
our inductive-recursive semantics, unlike in the set-theoretic model of
cumulativity for Coq \cite{timany18cumulative}, where function domains are
invariant.

@@

\item If levels are given by a limit ordinal, then every TTFL type is contained
in a universe. If levels form a successor ordinal, then this is not the
case. For example, Agda 2.6.1 has $\omega + 1$ levels (externally), where
$\Set{\omega}$ is the topmost universe, but $\Set{\omega}$ is not in any
universe.

@@

There have been proposals for strengthening universes with various closure
principles and universe operators. Palmgren's super universes and higher-order
universes \cite{Palmgren98onuniverses} and Setzer's Mahlo universes
\cite{setzer00mahlo} are examples for this. These are sufficient to model
transfinite hierarchies, but as we noted in Section
\ref{sec:inductive_recursive_codes}, we do not know how to model strict
inclusions with them. Variants of induction-recursion \cite{dybjer99finite,
  indexedir, positiveir} are particularly flexible and powerful extensions to
universes. McBride gave an inductive-recursion definition of cumulative
universes that we adapted in this work \cite{mcbride2015datatypes}.

@@

In the current work, we developed a framework for modeling a variety of universe
features in type theories. At this point, we may ask the question: if
induction-recursion is sufficient to model every feature, why not simply support
it in a practical implementation, and drop the menagerie of universe features?

The answer is that induction-recursion provides a \emph{deep embedding} of
universe features, which is usually less convenient to use than \emph{native}
features. For example, both Coq and Agda have powerful automatic solving for
filling out implicit universe levels. We also do not have to invoke $\El$ or the
$\ult$ computation rule explicitly, and in Coq we can use implicit syntax for
subtyping instead of explicit coercions.

@@

First, besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. We did keep
canonicity in mind when specifying the systems in this paper. Hopefully the
usual proof method of gluing (in other words, proof-relevant logical predicates)
\cite{coquandnorm,kaposi2019gluing,sterling2019algebraic} can be adapted to the theories in
this paper.

@@

Lastly, we do not handle impredicative universes. The main reason for this is
that we do not know the consistency of having induction-recursion and
impredicative function space together in the same universe, and modeling
impredicativity seems to require this assumption in the metatheory. This could
be investigated as well in future work.
